Characterization of gaps in directly bonded Si compound optics

### 2014-04-21 AO_bondedSi.tex
minor tweaks from MGS edits

While it is possible that proper precautions can be made to eliminate all micron-sized particles from both Si surfaces prior to bonding, the power-law particle size distribution in clean-room environments \cite{doi:10.1080/02786828608959094} means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as ``tent poles", that create gaps where the bonding is incomplete (Figure \ref{figParticle}).  Some authors \cite{1991JaJAP..30..615M, 1992JEMat..21..669M, 1989JaJAP..28L2141L, Mitani1990} have referred to these large ($\sim$ 1 $\mu$m size) gaps as \emph{IR bubbles}, or simply bubbles, owing to their appearance in infrared (IR) transmission images as contiguous regions of a deficit of IR transmission.  The gaps have also been referred to as voids\cite{2000RScI...71.1869G}, interfacial voids, or gaps.  In this paper we will reserve the term ``IR bubbles'' for gaps that are large enough to be detected directly in conventional IR imaging.  The term ``gap'' will apply more broadly to any finite size gap, exceeding the scale over which interfacial molecular forces are acting, roughly $>$2 nm.  We omit the adjective ``interfacial'' before gap for simplicity.

@@

Whatever technique one uses to remove or reduce the effects of gaps, it is essential to have a quantitative method by which to assess accurately the optical impact of the gaps before and after treatment.  This paper introduces a technique for directly measuring gaps at silicon-silicon interfaces.  Mitani and G\"oselle review other techniques for measuring and mapping gaps\cite{1992JEMat..21..669M}.  One key technique is IR Newton rings from IR imaging\cite{Mitani1990}.  This paper gives deeper understanding of the infrared interference technique seen in IR images as Newton rings.  The number and morphology of Newton rings yields information about the lateral and axial extent of large gaps. Specifically, dark-light fringes manifest when the gap is comparable to the vacuum wavelength or larger.  Our technique is IR interference in the limit of small gaps down to 20 nm or less.

There are other techniques for mapping gaps.  X-ray topography \cite{1992JEMat..21..669M,1994JaJAP..33....6H} exploits the change in optical density along lines of sight with gaps to allow precise and high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but provides only coarse estimates of the axial extent.  In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

@@

The foundation paper on Si bonding came in 1986\cite{1986JAP....60.2987S}.  After that, there was a sequence of papers using IR imaging to look for evidence of defects in the bond.  \cite{1988JaJAP..27L2364S, 1995ApPhL..67.3614G} directly monitored the bond propagation with IR imaging.  \cite{1988JaJAP..27L2364S, 1989JaJAP..28L2141L} demonstrated gap-free bonding in a micro-cleanroom, in which the bonding process occurs immediately after a rinse-spin-vacuum cycle inside of a vacuum enclosure.  Our group has not attempted the micro-cleanroom technique owing to the challenge of spin-cleaning large (30 mm thick) Si substrates.  It is not clear whether the custom micro-cleanroom technique has gained wide usage.  Notably, the   \cite{1989JaJAP..28L2141L} directly measured the bond surface energy.  \cite{Mitani1990} discussed the causes and remedies for bubble formation.  Specifically, the IR bubble density changes as a function of anneal temperature\cite{1992JEMat..21..669M}, with bubbles disappearing at temperatures greater than 1100$^\circ$ C.  The chemistry of the bonding interface began to be understood with infrared absorption spectroscopy and multiple internal reflection spectroscopy\cite{feijoo1994}, which lent credibility to the idea of distinct temperature phases and evidence for Si$-$H bonds\cite{1995ApPhA..61..101R}.  A robust picture of the chemistry as a function of anneal temperature was ultimately fleshed out\cite{1996JaJAP..35.2102R, 1998AnRMS..28..215G}.  Other variables potentially affecting bond strength were studied, for example surface roughness\cite{JJAP.37.4197}, surface topography \cite{2001JOptA...3...85G}, surface preparation\cite{1996ApPhL..68.2222T}, annealing time \cite{2000JAP....88.4404H}, ambient pressure and substrate thickness \cite{1995ApPhL..67..863G, 2007ApOpt..46.6793H}, and moisture and defects \cite{2001JAP....89.6013L}.  Notably, \cite{JJAP.37.4197} found that surface roughness above 1.3 nm RMS results in poor bonding quality in their sample of wafers etched with Argon.  Results from studies of surface topography are inconclusive\cite{2001JOptA...3...85G}.  Nowhere in the literature did we find a technique similar to the one presented here, probably because the technique relies on the high precision of relatively new IR capable spectroscopy that wasn't widely available until this decade.

@@

Consider the following two conceivable pitfalls in Si-Si direct bonding.  The obvious pitfall is particle contamination.  Figure \ref{figParticle} shows a schematic of a particle wedged between two flat Si optical surfaces.  A single 0.6 $\mu$m particle can create a dark IR bubble from interference.  In the next section we show that small gaps can unacceptably degrade the performance to as low as 35 nm.  In a class 100 cleanroom, the volume density could be as high as $\rho=100$ particles/ft$^3$, with sizes larger than 0.5 $\mu$m.  What we care about is not the volume number density of particles, but the delivered surface flux $F=\rho(D) v(D)$, where $v(D)$ is the average freefall speed of particles, which is a function of particle size.  Figure 5 of Cooper 1986\cite{doi:10.1080/02786828608959094} shows the surface flux (particles cm$^{-2}$ s$^{-1}$) vs. particle size ($\mu$m).  For a change from 1 $\mu$m to 10$^{-2}$ $\mu$m, there is an increase in a factor of $10^4$ in the particle flux expected.  In other words:
$$\log(F) = -2\log(D) + const.$$

So there will be roughly $10^4$ times more 10 nm sized particles than 1 $\mu$m sized particles.  For every obviously detectable IR bubble, there are thousands more particles that could unacceptably degrade the transmission performance.  One key idea that saves the day, is the intuition that the areal fill factor scales at least as fast as $D^2$.  Imagine that the particles serve as miniature tentpoles.  The volume beneath the tentpole looks like a cone.  Since the number density goes down as roughly $D^{-2}$, (per the equation above) the total area covered by particles might be constant in $log(D)$.  Simply put, the big particles cover just as much area as the small particles, even though there are many more small particles.  It is further conceivable that the areal fill factor attributable to a tent pole particle scales even faster than $D^2$ since surface energy probably scales inversely with distance of the pieces (\`a la VanDerWaals).  The small particles are even less of a problem than the large particles, since the small ($\sim10-60$ nm) particles will have lower average etalon reflectivity than larger (500-1000 nm) particles.  This last point will become clearer in later sections.

It has been shown\cite{1992JEMat..21..669M} that the gases seep through the Si$-$Si interface.  Pressure builds around catalysts.  In this sense, the size of the particle causing a gap cannot be uniquely identified from observations of the extent of gaps.  In section \ref{secResults} we address the question of the origin of interfacial gaps.  Specifically, are the particles more like \emph{tentpoles}, rigid structures that merely prevent the zippering up of the bond?  Or are the particles more like \emph{catalysts} so that their size is immaterial- the particles merely accelerate the production of gases that bow out the bond?  These are open questions.

\subsection{Surface roughness or large-scale non-conformance}
Another possible source of gaps at the bond is inherent roughness in the initial surfaces that leads to a failure to conform.  The second pitfall to consider is that the stiff non-flat Si surfaces will not conform upon contact leaving the complement of the aberrations as an air gap. Unconventional thick ($\sim$ 30 mm) Si applications a not as conformal as wafers, so we expect thick substrates could have gaps even in the absence of particles.  Most of our thick substrates are optically polished to much less than $\lambda = 632$ nm peak to valley over the 3 to 4 inch diameter clear aperture.

@@

Our measurement technique exploits the large Fresnel reflection \cite{2001opt4.book.....H} at the Si$-$vacuum interface.  The refractive index of silicon at room temperature ranges from 3.55$-$3.45 from $\lambda = $ 1150$-$2500 nm \cite{2006SPIE.6273E..77F}, one of the largest refractive indices of any conventional material.  The Fresnel reflection is accordingly about 30\% per surface.  For comparison, conventional optical materials with refractive indices in the range 1.5$-$1.7 have Fresnel reflection of merely 4$-$7\% per surface.  Figure \ref{figSiIndexFinesse} shows the refractive index as a function of wavelength for room temperature, which we computed using the coefficients from \cite{2006SPIE.6273E..77F}.

@@

The Si$-$Si interface gap is a low-Finesse Fabry-P\`{e}rot etalon\cite{2007fuph.book.....S}.  The transmission through a Fabry-P\`{e}rot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We assume transmission is at normal incidence, the refractive index of the gap is 1.0, and the material is at room temperature.  Si is negligibly absorbing longward of $\lambda$ = 1150 nm, which we verified with comparison of Si reference samples of different thicknesses.  We compute the reflectivity for an Si$-$vacuum interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \cite{2006SPIE.6273E..77F}, as shown in Figure \ref{figSiIndexFinesse}.  In the equations ahead we drop the subscripts on the refractive index $n(\lambda, T) \rightarrow n$ for clarity.  The computation is simply Fresnel's law at normal incidence:

@@

\subsection{Predicted transmission spectrum for bonded Si with finite interface gap}
\label{secTheory}
There is a key hurdle in computing the transmission spectrum for bonded Si with a finite interface gap- multiple incoherent reflections occur within the Si substrate at normal incidence.  Treatments for multiple coherent reflections have been described in detail in the optics literature \cite{2007fuph.book.....S}.  The wave transfer matrix technique treats each dielectric interface as a matrix with elements relating the pre- and post- interface complex amplitudes in the left and right directions.  We adapted the wave transfer technique for incoherent interactions \cite{2002ApOpt..41.3978K}.  Specifically I constructed an incoherent wave transfer matrix whose elements relate the intensities (and not complex amplitudes) before and after an interface.  The details are worked out in Appendix \ref{sec:Append-IMRTMM}.  We computed the transmission for two scenarios- first a double side polished (DSP) Si puck with no air gap, and second a pair of bonded Si pucks with a gap thickness $d$.  The DSP Si puck with no air gap has a transmission equal to:

$$
T_{DSP} = \frac{2n}{1+n^2} \label{eqnAbsDSPtrans}
$$

@@

$$T_{e} = \frac{n^2+1}{2 n F \sin ^2(2\pi \frac{d}{\lambda})+n^2+1} \label{eqFP}$$

>>>>

While it is possible that proper precautions can be made to eliminate all micron-sized particles from both Si surfaces prior to bonding, the power-law particle size distribution in clean-room environments \cite{doi:10.1080/02786828608959094} means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as ``tent poles", that create gaps where the bonding is incomplete (Figure \ref{figParticle}).  Some authors \cite{1991JaJAP..30..615M, 1992JEMat..21..669M, 1989JaJAP..28L2141L, Mitani1990} have referred to large ($\sim$ 1 $\mu$m size) gaps as \emph{IR bubbles}, or simply bubbles, owing to their appearance in infrared (IR) transmission images as contiguous regions of a deficit of IR transmission.  The gaps have also been referred to as voids\cite{2000RScI...71.1869G}, interfacial voids, or gaps.  In this paper we will reserve the term ``IR bubbles'' for gaps that are large enough to be detected directly in conventional IR imaging.  The term ``gap'' will apply more broadly to any finite size gap.

@@

Whatever technique one uses to remove or reduce the effects of gaps, it is essential to have a quantitative method by which to assess accurately the optical impact of the gaps before and after treatment.  This paper introduces a technique for directly measuring gaps at silicon-silicon interfaces.  Mitani and G\"oselle review other techniques for measuring and mapping gaps\cite{1992JEMat..21..669M}.  One key technique is identifying IR Newton rings from IR imaging\cite{Mitani1990}.  The number and morphology of Newton rings yields information about the lateral and axial extent of large gaps. Specifically, dark-light fringes manifest when the gap is comparable to the vacuum wavelength or larger.  Our technique is IR interference in the limit of small gaps down to 20 nm or less.

There are other techniques for mapping gaps.  X-ray topography \cite{1992JEMat..21..669M,1994JaJAP..33....6H} exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers.  In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

@@

The foundation paper on Si bonding came in 1986\cite{1986JAP....60.2987S}.  After that, there was a sequence of papers using IR imaging to look for evidence of defects in the bond.  \cite{1988JaJAP..27L2364S, 1995ApPhL..67.3614G} directly monitored the bond propagation with IR imaging.  \cite{1988JaJAP..27L2364S, 1989JaJAP..28L2141L} demonstrated gap-free bonding in a micro-cleanroom, in which the bonding process occurs immediately after a rinse-spin-vacuum cycle inside of a vacuum enclosure.  Our group has not attempted the micro-cleanroom technique owing to the challenge of spin-cleaning large (30 mm thick) Si substrates.  It is not clear whether the custom micro-cleanroom technique has gained wide usage.  \cite{1989JaJAP..28L2141L} directly measured the bond surface energy.  \cite{Mitani1990} discussed the causes and remedies for bubble formation.  Specifically, the IR bubble density changes as a function of anneal temperature\cite{1992JEMat..21..669M}, with bubbles disappearing at temperatures greater than 1100$^\circ$ C.  The chemistry of the bonding interface began to be understood with infrared absorption spectroscopy and multiple internal reflection spectroscopy\cite{feijoo1994}, which lent credibility to the idea of distinct temperature phases and evidence for Si$-$H bonds\cite{1995ApPhA..61..101R}.  A robust picture of the chemistry as a function of anneal temperature was ultimately fleshed out\cite{1996JaJAP..35.2102R, 1998AnRMS..28..215G}.  Other variables potentially affecting bond strength were studied, for example surface roughness\cite{JJAP.37.4197}, surface topography \cite{2001JOptA...3...85G}, surface preparation\cite{1996ApPhL..68.2222T}, annealing time \cite{2000JAP....88.4404H}, ambient pressure and substrate thickness \cite{1995ApPhL..67..863G, 2007ApOpt..46.6793H}, and moisture and defects \cite{2001JAP....89.6013L}.  Notably, \cite{JJAP.37.4197} found that surface roughness above 1.3 nm RMS results in poor bonding quality in their sample of wafers etched with Argon.  Results from studies of surface topography are inconclusive\cite{2001JOptA...3...85G}.  Nowhere in the literature did we find a technique similar to the one presented here, probably because the technique relies on the high precision of relatively new IR capable spectroscopy that was not widely available until this decade.

@@

There are two major conceivable pitfalls in Si-Si direct bonding.  The obvious pitfall is particle contamination.  Figure \ref{figParticle} shows a schematic of a particle wedged between two flat Si optical surfaces.  A single 0.6 $\mu$m particle can create a dark IR bubble from interference.  In a class 100 cleanroom, the particle volume density could be as high as $\rho=100$ particles/ft$^3$, for particle sizes larger than 0.5 $\mu$m.  What we care about is not the volume number density of particles, but the delivered surface flux $F=\rho(D) v(D)$, where $v(D)$ is the average freefall speed of particles, which is a function of particle size.  Figure 5 of Cooper 1986\cite{doi:10.1080/02786828608959094} shows the surface flux (particles cm$^{-2}$ s$^{-1}$) vs. particle size ($\mu$m).  There will be roughly $10^4$ times more 10 nm sized particles than 1 $\mu$m sized particles.  In other words:
$$\log(F) = -2\log(D) + const. \label{eqnParticleFlux}$$

For every obviously detectable IR bubble, there are thousands more particles that could unacceptably degrade the transmission performance.  One key idea that saves the day, is the intuition that the areal fill factor scales at least as fast as $D^2$.  Imagine that the particles serve as miniature tentpoles.  The volume beneath the tentpole looks like a cone.  Since the number density scales as roughly $D^{-2}$, (per the Equation \ref{eqnParticleFlux} above) the total area covered by particles might be constant in $log(D)$.  Simply put, the big particles cover just as much area as the small particles, even though there are many more small particles.  It is further conceivable that the areal fill factor attributable to a tent pole particle scales even faster than $D^2$ since surface energy probably scales inversely with distance of the pieces (\`a la VanDerWaals).  The small particles are even less of a problem than the large particles, since the small ($\sim10-60$ nm) particles will have lower average etalon reflectivity than larger (500-1000 nm) particles.  This last point will become clearer in later sections.

It has been shown\cite{1992JEMat..21..669M} that gases produced at the Si$-$Si interface seep along the interface surfaces.  Pressure builds around catalysts.  In this sense, the size of the particle causing a gap cannot be uniquely identified from observations of the extent of gaps.  In Section \ref{secResults} we address the question of the origin of gaps.  Specifically, are the particles more like \emph{tentpoles}, rigid structures that merely prevent the zippering up of the bond?  Or are the particles more like \emph{catalysts} so that their size is immaterial- the particles merely accelerate the production of gases that bow out the bond?  These are open questions.

\subsection{Surface roughness or large-scale non-conformance}
Another possible source of gaps at the bond is inherent roughness in the initial surfaces that leads to a failure to conform.  Stiff non-flat Si surfaces will not conform upon contact leaving the complement of the aberrations as an air gap. Unconventional thick ($\sim$ 30 mm) Si pucks are less conformal than wafers are, so we expect thick substrates could have gaps even in the absence of particles.  Most of our thick substrates are optically polished to less than 60 nm peak to valley over the 3 to 4 inch diameter clear aperture.  We assume that if enough pressure is placed on the substrates during bonding, then the surfaces will come in contact and zipper up.  Subsequent annealing will strength the bond.

@@

Our measurement technique exploits the large Fresnel reflection \cite{2001opt4.book.....H} at the Si$-$vacuum interface.  The refractive index of silicon at room temperature ranges from 3.55$-$3.45 from $\lambda = $ 1150$-$2500 nm \cite{2006SPIE.6273E..77F}, one of the largest refractive indices of any conventional material.  The Fresnel reflection is about 30\% per surface.  For comparison, conventional optical materials with refractive indices in the range 1.5$-$1.7 have Fresnel reflection of merely 4$-$7\% per surface.  Figure \ref{figSiIndexFinesse} shows the refractive index as a function of wavelength for room temperature, which we computed using the coefficients from \cite{2006SPIE.6273E..77F}.

@@

The Si$-$Si interface gap is a low-Finesse Fabry-P\`{e}rot etalon\cite{2007fuph.book.....S}.  The transmission through a Fabry-P\`{e}rot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We assume transmission is at normal incidence, the refractive index of the gap is 1.0, and the material is at room temperature.  Si is negligibly absorbing longward of $\lambda$ = 1150 nm, which we verified with comparison of Si reference samples of different thicknesses.  We compute the reflectivity for an Si$-$vacuum interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \cite{2006SPIE.6273E..77F}, as shown in Figure \ref{figSiIndexFinesse}.  In the equations ahead we drop the subscripts on the refractive index $n(\lambda, T) \rightarrow n$ and the coefficient of finesse $F(\lambda, T) \rightarrow F$ for clarity.  The computation is simply Fresnel's law at normal incidence:

@@

\subsection{Predicted transmission spectrum for bonded Si with finite interface gap}
\label{secTheory}
There is a key hurdle in computing the transmission spectrum for bonded Si with a finite interface gap- multiple incoherent reflections occur within the Si substrate at normal incidence.  Treatments for multiple coherent reflections have been described in detail in the optics literature \cite{2007fuph.book.....S}.  The wave transfer matrix technique treats each dielectric interface as a matrix with elements relating the pre- and post- interface complex amplitudes in the left and right directions.  We adapted the wave transfer technique for incoherent interactions \cite{2002ApOpt..41.3978K}.  Specifically I constructed an incoherent wave transfer matrix whose elements relate the intensities (and not complex amplitudes) before and after an interface.  The details are worked out in Appendix \ref{sec:Append-IMRTMM}.  We computed the transmission for two scenarios- first a double side polished (DSP) Si puck with no gap, and second a pair of bonded Si pucks with a gap thickness $d$.  The DSP Si puck with no air gap has a transmission equal to:

$$
T_{DSP} = \frac{2n}{1+n^2} \label{eqnAbsDSPtrans} \\
$$

@@

$$T_{e} = \frac{n^2+1}{2 n F \sin ^2(2\pi \frac{d}{\lambda})+n^2+1} \label{eqFP} \\$$
